Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-09-09 docs/Manuscript/Manuscript.tex

  \begin{abstract}
  Background: Applications of classical case-based reasoning (CBR) to
  clinical and health tasks have given rise to a family of techniques in
  which an (often conventional) model is fitted to a subset of labeled
  cases, selected for relevance or similarity to an unlabeled case to
  which the model is applied. These localized models, as we propose to
  call them, may perform as well as ``black-box'' models while retaining
  direct interpretations. Their adoption has been hindered by fragmented
  development and implementation challenges. Our goal in this scoping
  review was to describe the clinical and health applications of
  localized models to date and to derive a general framework for their
  design and evaluation.

@@

  Results: Of 1,309 search results, 328 were reviewed. Of these,
  combined with 43 publications that seeded the review and 1 obtained by
  citation tracking, 25 were included in the review and methodological
  synthesis. The specificity of several search terms was poor, and
  inter-rater reliability was low. Entries were published from 1997 to
  2021, with half of these since 2015. The most common tasks were
  prognosis and diagnosis, with only few recent studies focusing on care
  recommendation. The studies overwhelmingly focused on clinical data,
  occasionally also including laboratory and image-derived data. Only
  three studies used localized models for any task other than
  prediction, most recently to identify significant determinants behind
  recommendation outputs. Several studies commented on memory and
  runtime costs, and four proposed partial solutions to them. While most
  techniques were not replicated and did not cite predecessors, a
  general technique that specializes to most of those encountered was
  straightforward to describe.

  Conclusions: Localized models hold promise to improve the performance
  of models used in clinical decision support tools while maintaining
  interpretability, provided their computational limitations can be
  overcome. Our search undoubtedly failed to capture all studies that
  used methods of this kind, but those reviewed and synthesized exhibit
  consistent enough design to be reproduced by a general-purpose tool.
  We hope that our review and synthesis will assist the development of
  more efficient implementations suitable for integration into practice.
  \end{abstract}
    \begin{keyword}
    case-based reasoning \sep localized modeling \sep nearest
neighbors \sep clinical decision support \sep 
    scoping review
  \end{keyword}

@@

Clinical decision support (CDS) systems have been a fixture of
artificial intelligence (AI) research since inception. As machine
learning (ML) emerged, clinical modeling tasks became some of the most
common and celebrated applications. A technique core to clinical
applications of both AI and ML is \emph{similarity-matching}, which
relies on a \emph{patient similarity measure}\footnote{Similarity
  measures between more granular units of analysis, such as encounters
  and decision points, are often still referred to as patient similarity
  measures.} to extract a cohort of past or training cases used to
inform the diagnosis, prognosis, or care of a new or test case.
Similarity-matching powers the retrieval step of most case-based
reasoning (CBR) systems and provides the neighborhoods from which
nearest neighbor (NN) models derive predictions. Whereas NN predictions
are generated automatically via averaging (regression) or voting
(classification), traditional CBR tools return the retrieved cases to
the user to inform human decision-making.

We discuss in this review a hybrid computational approach we call
\emph{localized modeling} that borrows elements from both clinical
application areas and appears to have been introduced independently
several times. The approach fits ML models, including regularized
families of generalized regression models (GLMs), to cohorts retrieved
via similarity matching. Whereas CBR and NN prediction use patient
similarity to obtain an ideally homogeneous subpopulation matched to the
index case, in localized modeling one seeks a relevant yet heterogeneous
subpopulation, each member of which deviates slightly from the index
case. In principle, this allows interpretable models like GLMs to
identify and measure the effects of determining factors that are
specific to the index case. The use of similarity cohorts distinguishes
localized modeling from classical GLMs that still dominate analysis in
clinical trials and comparative effectiveness research, as well as from
many cutting-edge ML models that are fitted to whole populations, while
the cohorts' heterogeneity and its utility to the model distinguish the
approach from CBR. In the reviewed studies, localized modeling shows
some promise toward simultaneously achieving the performance
improvements of ML and the interpretability of classical models.

@@

Clinical CBR emerged among rule-based approaches and other AI tools in
the development of expert systems {[}2{]}. Early implementations
realized a general workflow described as ``the four REs'', later the
``R4 cycle'' {[}2,3{]}: Given a new case (or problem), the system
retrieves one or more past cases (solved problems) from a corpus, reuses
these to generate an understanding of (solution to) the new case,
revises this understanding to better fit the new case (sometimes called
``adaptation''), and retains the new case and its eventual understanding
in the corpus to be retrieved and reused in future. In contrast to
rule-based systems, which are variable-based and generally interpretable
as a single rule applied to all new cases, case-based systems provide
case-specific interpretations in the form of a number of more fully
understood reference cases. While the similarity measure used in the
retrieval step need not be changed as the corpus grows, proposed
measures have been diverse, contested, and rarely systematically
compared.

{[}4{]} distinguished two styles of CBR: problem-solving, which is more
procedural and used when objectives are more clearly defined, and
interpretative, which provides categorizations and justifications for
possible solutions. A similar dichotomy is commonly used to distinguish
the performance criterion for the usefulness of predictive models from
the interpretability criterion for their usability. In these terms,
localized models are highly procedural: The step of fitting a predictive
model to a retrieved cohort can be understood as an automated adaptive
strategy {[}3{]}, and the parameters that govern cohort retrieval can be
tuned alongside model hyperparameters in a ML workflow. Accordingly,
localized models have been primarily used for and evaluated on their
ability to more accurately predict classes or outcomes. Nevertheless, as
some more recent applications have begun to show, they can be used to
draw inferences about the importance of different risk factors to
specific individuals. \textbf{Distinguish between interpretable model
components like effect estimates and informative but not directly
interpretable importance measures as are commonly computed for
`black-box' models.} While the patient similarity measure used to
retrieve each cohort may be complicated, the cohort itself can be
directly inspected by the user. Provided the model family fitted to the
cohorts is interpretable, the localized model as a whole inherits this
property. Thus, localized models may embody both styles of CBR.

We refer the interested reader to several previous reviews of CBR in
medicine {[}3,5,6{]}, of measures of patient similarity {[}7{]}, and of
uses of patient similarity in predictive models {[}8--10{]}. While the
reviews of CBR focus on applications using health data, the patient
similarity reviews encompass many additional types of data (various
molecular -omics, genetic tests, medical images, laboratory tests,
patient preferences, patient-reported outcomes, tracking devices, social
media) and survey a much broader scope of models (exploratory analysis
via dimension reduction and cluster analysis; risk evaluation and
outcome prediction; clinical decision support and software tools). For
the present review, we are interested in how similarity matching on
patient-level health data is used to construct local cohorts for
predictive modeling.

@@

\begin{itemize}
\tightlist
\item
  Pulls case-level data from a corpus of past cases with known response
  values\newline (response may be outcome, diagnosis, subtype, etc.)
\item
  Defines a numeric multivariate case similarity measure\newline
  (include integer-valued measures)
\item
  Uses the similarity measure to select cohorts for index cases from the
  corpus\newline (for example, based on a training--testing partition)
\item
  Fits statistical models to cohorts from which to draw inferences about
  index cases\newline (outcome predictions count, survival estimates,
  risk factor effect estimates, model evaluation statistics, etc.)
\end{itemize}

We expected the general framework to boil down to three choices: A
patient similarity measure, a cohort selection process, and a
statistical model family. Such a general method would be
straightforward, if laborious, to implement, and specializations of it
would account for the majority of real-world applications. A general
implementation based on this method would allow researchers to expedite
every step of the analysis process, including selection, optimization,
and evaluation, and enable sensitivity, robustness, and multiverse
analyses that help identify the most consequential choices along the
way.

@@

We describe the steps of our review in detail in the Appendix. We also
note deviations from plans and the reasons for them.

@@

We include a PRISMA checklist as Supplemental Table 1 and a PRISMA
abstract checklist as Supplemental Table 2. Because we focus on
methodologies rather than approaches, we deviate in some ways from
PRISMA guidelines. In particular, many items of the checklist, involving
bias assessment and quantitative synthesis, are intended for
meta-analyses and, therefore, did not apply to this study. Because we
are not aware of any standard procedures for conducting reviews and
syntheses of methodology, no protocol was prepared for this study.

@@

Within Covidence, we screened titles and abstracts for relevance. We
decided on four eligibility criteria to expedite the screening process.
These include articles written in English for readability, original
studies to strictly include primary sources with new
information/conclusions, medical or clinical related settings due to the
applicability of our data to this field, and alternative use of the key
terms within the search string, meaning that it was clear from the title
and abstract that the entry was included due to the use of one or more
search terms with a different meaning than ours. Each of two authors (AC
and JCB) screened roughly half of the entries. They regularly reviewed
each other's decisions to improve consistency. In cases of uncertainty,
an entry would be included.

@@

@@

\begin{itemize}
\tightlist
\item
  Pulls case-level data from a corpus of past cases with known classes
  or outcomes
\item
  Defines a numeric multivariate case similarity measure
\item
  Uses the similarity measure to select cohorts for index cases from the
  corpus
\item
  Fits statistical models to cohorts to make inferences about index
  cases
\end{itemize}

@@

In Covidence, two authors (AC and JCB) independently evaluated each
manuscript for the four eligibility criteria. The first criterion that a
manuscript failed was designated the reason for exclusion. An antecedent
criterion was used to exclude manuscripts that did not report the
results of original studies involving real-world experiments or
empirical data, for example surveys of prior work and proposals of
frameworks. In cases of disagreement between the authors on the reason
for exclusion, the first criterion was adopted. The authors arrived at
agreement on inclusion or exclusion through discussion. We calculated
inter-rater reliability to evaluate our screening and review process.

During full text review, we decided to expand the conception of
statistical models (fourth criterion): Rather than restricting to models
that are fit and evaluated in separate steps, we chose to allow simple
statistical summaries such as mean survival {[}11{]} and survival curves
{[}12{]}. These approaches were novel to CBR and presage later
developments, so were helpful to understanding the development of
localized modeling. However, this then admitted studies that applied
conventional nearest neighbors prediction: Each retrieved cohort
consisted of the \(k\) most similar cases to the index case, and the
response for the index case was predicted to be either the mean
(continuous response) or the plurality (discrete response) of the
responses of the cohort. A review of all studies that use nearest
neighbors prediction, even in the clinical domain, would be impractical.
Because our focus is on novel approaches that combine similarity-based
retrieval and statistical modeling of retrieved cohorts, we chose to
exclude those studies whose approach was equivalent to nearest neighbors
prediction using a conventional similarity measure.

@@

Following the selection sources aligning with our review, we collected
characteristics of included studies (Section \ref{sec:data}). The
features included bibliographic fields (date of publication, journal,
authors, title, keywords, DOI), study goals (objective, generalizable
knowledge, evaluation, clinical/medical domain), data sets (data source,
type of data, range of data, number of cases/incidences, number of
predictors/features), and methodological choices (types of similarity
measure, families of adaptation step/statistical model, method(s)
compared against, evaluation/performance/comparison measures (\& their
values), name given to modeling approach). We used these data to detect
and visualize trends amongst the included studies.

@@

We then reviewed 43 studies comprising a seed set that inspired this
review. After removing duplicates and screening for eligibility, we were
left with 6 additional studies {[}12--17{]}, 1 of which was excluded
from the synthesis for using nearest neighbors prediction. Finally,
citations in reference lists from the 43 + 328 = 371 studies assessed
for eligibility led us to identify 1 additional study that met our
criteria {[}18{]}. This left us with 19 + 1 + 5 = 25 studies included in
the review and synthesis.

@@

@@

Another interesting pattern lay in the years of publication of the
included studies (Figure \ref{fig:year}). While they stretch back to the
late 1990s, they have become more common, which suggests that this is
still an active, though not rapidly expanding, methodological approach.
We note that, while CBR in health and medicine originated as early as
1990, these studies did emerge early on once CBR had established itself.
So the idea has been ``in the air'' for as long as CBR has been in use.

@@

Another dominant characteristic of included studies is their broader
remit. We categorized studies as ``knowledge'' or ``practice'' based on
whether their aim was to produce generalizable knowledge or to improve
practice. We classified 7 of the 25 studies as ``knowledge'', making the
majority of studies from a ``practice'' standpoint. These studies aimed
primarily to provide providers with tools to use in clinic or to improve
outcomes for patients currently receiving care.

@@

The included studies hypothesized, asserted, or assumed several benefits
of localized modeling specific to clinical and medical settings and
advantages over other modeling approaches (see Section \ref{sec:data}).
The most common was that the restriction to similar or relevant past
cases would improve predictive performance for the index case
{[}11,15,16,26{]}. In particular, {[}14{]} hypothesized and confirmed
that the value of each past case would be positively related to its
similarity to the index case, an assumption built in to the weighting
schemes of other approaches. {[}12{]} made a different case, that the
fewer parametric assumptions and complications of a CBR-style model
would allow for greater accuracy. Additionally, several investigators
asserted that the use of localized cohorts befit the clinical focus on
the individual patient rather than the population, without reference to
model performance {[}21,23{]}.

@@

The other frequent advantage attributed to localized modeling was
interpretability. {[}22{]} and {[}28{]} emphasized the value of the
detailed past cases, available to the user, on which predictions are
based. This ``self-explanation capability'' made outputs more
intelligible to physicians in the CDS setting. {[}15{]} additional
pointed out that localized GLMs yield localized effect estimates, which
may help investigators identify individually relevant risk factors.

@@

CBR can be understood as the opposite side of a trade-off with
rule-based reasoning between model size and model complexity; {[}36{]}
describe their approach as ``knowledge-light'', in that ``the cases do
not contain explicit explanation structures; instead, explanation is
achieved by comparison of the query case with retrieved cases''. This
means that the greatest performance and efficiency challenges in CBR
have to do with the retrieval and revision phases in the R4 cycle.
Several studies addressed these challenges: {[}13{]}, while excluded
from the synthesis, were the earliest to propose that cohorts be bounded
by a similarity threshold rather than by a number of cases, and this
improved performance in their experiments. {[}38{]} leveraged predictor
weights obtained from logistic regression to inform the similarity
calculation used in retrieval. {[}12{]} used non-parametric models to
reduce the computational burden of revision (prediction). {[}32{]}
proposed to improve efficiency along the entire R4 cycle, but in
particular by only executing task-dependent steps in real time
(``just-in-time learning''), and {[}35{]} and {[}34{]} partitioned
multiple phases in the R4 cycle into offline and online components, only
the latter of which would be performed in real time as new data are
received.

Other studies addressed limitations of available tools. {[}24{]}
implemented a comprehensive CBR tool in response to the lack of
general-purpose software, to enable coupling with other tools as well as
to expedite development, experimentation, and uptake. Several other
teams also set out to develop more interoperable clinical support tools
{[}15,22,26,30{]}.

The frequency plot in Figure \ref{fig:methods} compares the rates at
which several methodological elements are invoked in the sample.

@@

However, we do observe one clear pattern: Of the proposed methods that
most evidently outperformed their comparators (Kohonen + CBR {[}20{]},
TWNFI {[}18{]}, GSA {[}26{]}, CBR + rules (+ DML) {[}28{]}, Gaussian
process regression {[}30{]}, JITL-ELM {[}32{]}), all except the last are
hybrids of localized modeling (in some cases CBR) with other techniques,
often metric learning. Though {[}27{]} and {[}15{]} report exceptions,
the pattern suggests the importance of the similarity measure to the
retrieval step. Meanwhile, when proposed approaches targeted cohort
demarcation or choice of predictive model (statistical CBR {[}13{]};
individualized logistic regression, decision tree, and random forest
{[}14,16{]}), they did not consistently outperform comparators.

@@

The study authors focused their recommendations and their own plans for
future work mostly on technical improvements and evaluations
{[}11,19{]}. Urged improvements included full or partial automation of
user-dependent steps such as predictor selection {[}11,19{]}, similarity
learning {[}11,17{]}, and parameter optimization {[}16,21{]}; extensions
to new data structures {[}24{]}, data types {[}25,26,31{]}, and
reasoning systems {[}28{]}; and the use of more advanced model
components to improve accuracy or efficiency {[}12,14,26,30{]}. Authors
also urged validations and independent evaluations using larger or more
comprehensive data sets {[}15,22,23,25{]}, using data aggregated from
multiple health systems {[}14,16,34,35{]}, and in other care settings or
disease contexts {[}21,30,34,35{]}. Less common were calls to strengthen
the connection between the methods and the users. Some authors urged the
incorporation of (human-derived) domain knowledge into the data or
models {[}17,19{]}. Others suggested new uses of their modeling
approaches: to measure feature importance {[}20{]}, to make models more
expressive {[}14{]}, and to combine information obtained both from
global and from localized models {[}15{]}. Most of this incremental work
was indeed carried out in later investigations in our sample.

\hypertarget{public-availability}{%
\subsection{Public availability}\label{public-availability}}

\label{sec:availability}

@@

As discussed in Section \ref{sec:rationales}, most studies were premised
on the potential predictive value of localized models. As cautioned in
Section \ref{sec:performance}, not all experiments affirmed this
premise. Most of those that did involved using metric learning to
improve retrieval and sometimes iterative optimization of the metric and
of the localized model. Some studies showcased results using a variety
of such specializations {[}15,27,30{]}. However, because these advanced
implementations were mostly compared against their precursors or
commonplace alternatives, we cannot speak to their performance or any
trade-offs with respect to each other. This would be a worthy goal of
future work, though it might be limited by the lack of open-source code
or public implementations (Section \ref{sec:availability}).

@@

\textbf{Write this section!}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Deduplication (Covidence)

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    Specific to that source: title, creator, URL, DOI
  \item
    May overlap: journal, publication date, ISSN
  \end{enumerate}
\item
  Title

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    Exclude non clinical/non medical setting
  \item
    Must be in English
  \item
    Must be original study (not reviews, surveys, opinion, news)
  \item
    Exclude if search term clearly has different meaning than intended
  \end{enumerate}
\item
  Abstract
\item
  Full Text

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    Study criteria (inclusion/exclusion)
  \end{enumerate}
\end{enumerate}

@@

>>>>

  \begin{abstract}
  Background: Applications of classical case-based reasoning (CBR) to
  clinical and health tasks have given rise to a family of techniques in
  which a statistical model is fitted to a subset of labeled cases,
  selected for relevance or similarity to an unlabeled case to which the
  model is applied. These localized models, as we propose to call them,
  may perform as well as ``black-box'' models while retaining direct
  interpretations. Their adoption has been hindered by fragmented
  development and implementation challenges. Our goal in this scoping
  review was to describe the clinical and health applications of
  localized models to date and to derive a general framework for their
  design and evaluation.

@@

  Results: Of 1,309 search results, 328 were reviewed. Of these,
  combined with 43 publications that seeded the review and 1 obtained by
  citation tracking, 25 were included in the review and methodological
  synthesis. The specificity of several search terms was poor, and
  inter-rater reliability was low. Entries were published from 1997 to
  2021, with half of these since 2015. The most common tasks were
  prognosis and diagnosis. The studies overwhelmingly focused on
  clinical data, occasionally including laboratory and image-derived
  data. Only three studies applied localized models to tasks other than
  prediction. Several studies commented on memory and runtime costs, and
  four proposed partial solutions. While few methods were reused and no
  results were reproduced, a general technique that specializes to most
  of those encountered was straightforward to describe.

  Conclusions: Localized models have potential to improve the
  performance of models used in clinical decision support tools while
  maintaining interpretability, provided computational hurdles can be
  overcome. Our search undoubtedly failed to capture all studies that
  used methods of this kind, but those we found follow a consistent
  enough design to be reproduced by a general-purpose tool. We hope that
  our review and synthesis will spur future work on efficiency,
  reproducibility, and user interactions and needs.
  \end{abstract}
    \begin{keyword}
    case-based reasoning \sep localized modeling \sep nearest
neighbors \sep clinical decision support \sep 
    scoping review
  \end{keyword}

@@

Clinical decision support (CDS) systems have been a fixture of
artificial intelligence (AI) research since inception. As machine
learning (ML) emerged, clinical modeling tasks became some of the most
common and celebrated applications. A technique core to clinical
applications of both AI and ML is \emph{similarity-matching}, which
relies on a \emph{patient similarity measure}\footnote{Similarity
  measures between more granular units of analysis, such as encounters
  and decision points, are often still referred to as patient similarity
  measures.} to extract a cohort of past or training cases used to
inform the diagnosis, prognosis, or care of a new or test case.
Similarity-matching powers the retrieval step of most case-based
reasoning (CBR) systems and provides the neighborhoods from which
nearest neighbor (NN) models derive predictions. Whereas NN predictions
are generated automatically via averaging (regression) or voting
(classification), traditional CBR tools return the retrieved cases to
inform human decision-making.

We discuss in this review a hybrid computational approach we call
\emph{localized modeling} that appears to have been introduced
independently several times. The approach fits ML models, including
regularized families of generalized regression models (GLMs), to cohorts
retrieved via similarity matching. Whereas CBR and NN prediction use
patient similarity to obtain an ideally homogeneous subpopulation
matched to the index case, in localized modeling one seeks a relevant
yet heterogeneous subpopulation, each member of which may deviate from
the index case. A prediction or inference is then obtained from a model
fitted to this \emph{similarity cohort}. The use of similarity cohorts
distinguishes localized modeling from classical GLMs that still dominate
analysis in clinical trials and comparative effectiveness research, as
well as from many cutting-edge ML models that are fitted to whole
populations, while the cohorts' heterogeneity and its utility to the
model distinguish the approach from CBR. In principle, this allows
interpretable models like GLMs to identify and measure the effects of
determining factors that are specific to the index case. In the reviewed
studies, localized modeling shows some promise toward simultaneously
achieving the performance improvements of ML and the interpretability of
classical models.

@@

Clinical CBR emerged among rule-based approaches and other AI tools in
the development of expert systems {[}2{]}. Early implementations
realized a general workflow described as ``the four REs'', later the
``R4 cycle'' {[}2,3{]}: Given a new case (or problem), the system
\emph{retrieves} one or more past cases (solved problems) from a corpus,
\emph{reuses} these to generate an understanding of (solution to) the
new case, \emph{revises} this understanding (solution) to better fit the
new case (sometimes called ``adaptation''), and \emph{retains} the new
case and its eventual understanding (solution) in the corpus to be
retrieved and reused in future. In contrast to rule-based systems, which
are variable-based and generally interpretable as a single rule applied
to all new cases, case-based systems provide case-specific
interpretations in the form of a number of more fully understood
reference cases. While the similarity measure used in the retrieval step
need not be changed as the corpus grows, proposed measures have been
diverse, contested, and rarely systematically compared.

{[}4{]} distinguished two styles of CBR: problem-solving, which is more
procedural and used when objectives are more clearly defined, and
interpretative, which provides categorizations and justifications for
possible solutions. A similar dichotomy is commonly used to distinguish
the performance criterion for the usefulness of predictive models from
the interpretability criterion for their usability. In these terms,
localized models are highly procedural: The step of fitting a predictive
model to a retrieved cohort can be understood as an automated adaptive
strategy {[}3{]}, and the parameters that govern cohort retrieval can be
tuned alongside model hyperparameters in a ML workflow. Accordingly,
localized models have been primarily used for and evaluated on their
ability to predict classes or outcomes. Nevertheless, as some recent
applications have begun to show, they can be used to draw inferences
about the importance of different risk factors to specific individuals.
While most ML models come equipped with measures of feature importance
and model-agnostic tools can generate explanations for model
predictions, these quantifications are not directly interpretable model
components analogous to the split nodes of a decision tree (DT) or the
effect estimates of a GLM. While the patient similarity measure used to
retrieve each cohort may be complicated, the cohort itself can be
directly inspected by the user. Provided the model family fitted to the
cohorts is interpretable, the localized model as a whole inherits this
property. Thus, localized models may embody both styles of CBR.

We refer the interested reader to several previous reviews of CBR in
medicine {[}3,5,6{]}, of measures of patient similarity {[}7{]}, and of
uses of patient similarity in predictive models {[}8--10{]}. While the
reviews of CBR focus on applications using health data, the patient
similarity reviews encompass many additional types of data (various
molecular -omics, genetic tests, medical images, laboratory tests,
patient preferences, patient-reported outcomes, tracking devices, social
media) and survey a much broader scope of models (exploratory analysis
via dimension reduction and cluster analysis; risk evaluation and
outcome prediction; clinical decision support and software tools). For
the present review, we are interested in how similarity matching on
patient-level health data is used to construct local cohorts for
predictive or inferential modeling.

@@

\begin{itemize}
\tightlist
\item
  Uses case-level data from a corpus of past cases with known
  responses\newline (response may be outcome, diagnosis, subtype, etc.)
\item
  Defines a numeric multivariate case similarity measure\newline (allow
  integer-valued measures)
\item
  Uses the similarity measure to retrieve cohorts for index cases from
  the corpus\newline (for example, based on a training--testing
  partition)
\item
  Fits statistical models to cohorts from which to make predictions or
  draw inferences about index cases\newline (outcome predictions,
  survival estimates, risk factor contributions, model evaluation
  statistics, etc.)
\end{itemize}

We expected the general framework to come down to three choices: A
patient similarity measure, a cohort selection process, and a
statistical model family. Such a general method would be
straightforward, if laborious, to implement, and specializations of it
would account for the majority of real-world applications. A general
implementation based on this method would allow researchers to expedite
every step of the analysis process, including selection, optimization,
and evaluation, and enable sensitivity, robustness, and multiverse
analyses that help identify the most consequential choices along the
way.

@@

Here we describe our review process, including deviations from plans and
the reasons for them. More details are included in Section
\label{sec:appendix-selection}.

@@

\textbf{Correct ``n = PubMed'' in PRISMA diagram.} We include a PRISMA
checklist as Supplemental Table 1 and a PRISMA abstract checklist as
Supplemental Table 2. Because we focus on methodologies rather than
conceptual approaches or evidence, we deviate in some ways from PRISMA
guidelines. In particular, those items of the checklist involving bias
assessment and quantitative synthesis are intended for meta-analyses so
did not apply to this study. Because we are not aware of any standard
procedures for conducting reviews and syntheses of methodology, no
protocol was prepared for this study.

@@

Within Covidence, we screened titles and abstracts for relevance. We
decided on four eligibility criteria to expedite the screening process.
Articles must be written in English, for readability; they must be
original studies, to exclude secondary sources with duplicate
information; their use settings must be medical, clinical, or related,
to keep our review topical; and it must not be clear that their use of
our search terms was different from our intended meaning. Each of two
authors (AC and JCB) screened roughly half of the entries. They
regularly reviewed each other's decisions to improve consistency. In
cases of uncertainty, entries were included.

@@

\label{sec:full-text}

@@

\begin{itemize}
\tightlist
\item
  Pulls case-level data from a corpus of past cases with known classes
  or outcomes
\item
  Defines a numeric multivariate case similarity measure
\item
  Uses the similarity measure to select cohorts for index cases from the
  corpus
\item
  Fits statistical models to cohorts to make predictions or inferences
  about index cases
\end{itemize}

@@

In Covidence, two authors (AC and JCB) independently evaluated each
manuscript for these eligibility criteria. The first criterion that a
manuscript failed was designated the reason for exclusion. An antecedent
criterion was used to exclude manuscripts that did not report the
results of original studies involving real-world experiments or
empirical data, for example surveys of prior work and proposals of
frameworks. In cases of disagreement between the authors on the reason
for exclusion, the first criterion was adopted. The authors arrived at
agreement on inclusion or exclusion through discussion. We calculated
inter-rater reliability to evaluate our screening and review process.

During full text review, we decided to expand the conception of
statistical models (fourth criterion): Rather than restricting to models
that are fit and evaluated in separate steps, we chose to allow simple
statistical summaries such as mean survival {[}11{]} and survival curves
{[}12{]}. These approaches were novel to CBR and presage later
developments, so were helpful to understanding the development of
localized modeling. However, this then admitted studies that applied
conventional nearest neighbors prediction: Each retrieved cohort
consisted of the \(k\) most similar cases to the index case, and the
response for the index case was predicted to be either the mean
(continuous response) or the plurality (discrete response) of the
cohort's responses. A review of all studies that use nearest neighbors
prediction would be impractical. Because our focus is on novel
approaches that combine similarity-based retrieval and statistical
modeling of retrieved cohorts, we chose to exclude those studies whose
approach was equivalent to nearest neighbors prediction using a
conventional similarity measure.

@@

Following the selection sources aligning with our review, we collected
characteristics of included studies (Section \ref{sec:data}). The
features included bibliographic fields (date of publication, journal,
authors, title, keywords, DOI), study goals (objective, generalizable
knowledge, evaluation, clinical/medical domain), data sets (data source,
type of data, range of data, number of cases/incidences, number of
predictors/features), and methodological choices (types of similarity
measure, families of adaptation step/statistical model, method(s)
compared against, performance measures, results of evaluations and
comparisons, name given to modeling approach). We used these data to
detect and visualize trends amongst the included studies.

@@

We then reviewed 43 studies comprising a seed set that inspired this
review. After removing duplicates and screening for eligibility, we were
left with 6 additional studies {[}12--17{]}, 1 of which was excluded
from the synthesis for using NN prediction. Finally, reference tracking
from the 43 + 328 = 371 studies assessed for eligibility led us to
identify 1 additional study that met our criteria {[}18{]}. This left us
with 19 + 1 + 5 = 25 studies included in the review and synthesis.

@@

\textbf{Update numbers below, perhaps dynamically.}

@@

Another interesting pattern lay in the years of publication (Figure
\ref{fig:year}). While they trace back to the late 1990s, they have
become more common, which suggests that this is an active, though not
rapidly expanding, methodological approach. We note that, while CBR in
health and medicine originated as early as 1990, these studies did
emerge early on once CBR had established itself. So the idea has been
``in the air'' for as long as CBR has been in use.

@@

Another dominant characteristic of included studies is their broader
remit. We categorized studies as ``knowledge'' or ``practice'' based on
whether their aim was to produce generalizable knowledge or to improve
practice. We classified 7 of the 25 studies as ``knowledge'', making the
majority of studies from a ``practice'' standpoint. These studies aimed
to provide tools for use in clinic or to improve outcomes.

@@

The included studies hypothesized, asserted, or assumed several benefits
of localized modeling specific to clinical and medical settings and
advantages over other modeling approaches (see Section \ref{sec:data}).
The most common was that the restriction to similar or relevant past
cases would improve predictive performance for the index case
{[}11,15,16,26{]}. In particular, {[}14{]} hypothesized and confirmed
that the value of each past case would be positively related to its
similarity to the index case, an assumption built in to the weighting
schemes of other approaches. {[}12{]} made a different case, that the
fewer parametric assumptions and complications of a CBR-style model
would allow for greater accuracy. Additionally, several investigators
asserted that the use of localized cohorts befit the clinical focus on
the individual patient rather than the population, without reference to
performance {[}21,23{]}.

@@

The other frequent advantage attributed to localized modeling was
interpretability. {[}22{]} and {[}28{]} emphasized the value of the
detailed past cases, available to the user, on which predictions are
based. This ``self-explanation capability'' made outputs more
intelligible to physicians in the CDS setting. {[}15{]} additionally
pointed out that localized GLMs yield localized effect estimates, which
may help investigators identify individually relevant risk factors.

@@

CBR can be understood as the opposite side of a trade-off with
rule-based reasoning between model size and model complexity; {[}36{]}
describe their approach as ``knowledge-light'', in that ``the cases do
not contain explicit explanation structures; instead, explanation is
achieved by comparison of the query case with retrieved cases''. This
means that the greatest performance and efficiency challenges in CBR
have to do with the retrieval and revision phases in the R4 cycle.
Several studies addressed these challenges: {[}13{]}, while excluded
from the synthesis, were the earliest to propose that cohorts be bounded
by a similarity threshold rather than by a number of cases, and this
improved performance in their experiments. {[}38{]} leveraged predictor
weights obtained from logistic regression to inform the similarity
calculation used in retrieval. {[}12{]} used non-parametric models to
reduce the computational burden of revision (prediction). {[}32{]}
proposed to improve efficiency along the entire R4 cycle, but in
particular by only executing task-dependent steps in real time
(``just-in-time learning'', JITL), and {[}35{]} and {[}34{]} partitioned
multiple phases in the R4 cycle into offline and online components, only
the latter of which would be performed in real time as new data are
received.

Other studies addressed limitations of available tools. {[}24{]}
implemented a comprehensive CBR tool in response to the lack of
general-purpose software, to enable coupling with other tools as well as
to expedite development, experimentation, and uptake. Several other
teams also set out to develop more interoperable clinical support tools
{[}15,22,26,30{]}. The frequency plot in Figure \ref{fig:methods}
compares the rates at which several methodological elements are invoked
in the sample.

@@

We do observe one clear pattern: All of the proposed methods that most
evidently outperformed their comparators (Kohonen + CBR {[}20{]}, TWNFI
{[}18{]}, gravitational search algorithms (GSA) {[}26{]}, CBR + rules (+
DML) {[}28{]}, Gaussian process regression {[}30{]}, JITL-ELM {[}32{]})
are hybrids of localized modeling (in some cases CBR) with other
techniques, often metric learning. Though {[}27{]} and {[}15{]} report
non-superior performance by such hybrids, the pattern suggests the
importance of the similarity measure to the retrieval step. Meanwhile,
when proposed approaches targeted cohort demarcation or choice of
predictive model (statistical CBR {[}13{]}; individualized logistic
regression, decision tree, and random forest {[}14,16{]}), they did not
consistently outperform comparators.

@@

The study authors focused their recommendations and their own plans for
future work mostly on technical improvements and evaluations
{[}11,19{]}. Urged improvements included full or partial automation of
predictor selection {[}11,19{]}, similarity learning {[}11,17{]}, and
parameter optimization {[}16,21{]}; extensions to new data structures
{[}24{]}, data types {[}25,26,31{]}, and reasoning systems {[}28{]}; and
the use of more advanced model components to improve accuracy or
efficiency {[}12,14,26,30{]}. Authors also urged validations and
independent evaluations using larger or more comprehensive data sets
{[}15,22,23,25{]}, using data aggregated from multiple health systems
{[}14,16,34,35{]}, and in other care settings or disease contexts
{[}21,30,34,35{]}.

Less common were calls to strengthen the connection between the methods
and the users. Some authors urged the incorporation of (human-derived)
domain knowledge into the data or models {[}17,19{]}. Others suggested
new uses of their modeling approaches: to measure feature importance
{[}20{]}, to make models more expressive {[}14{]}, and to combine
information obtained both from global and from localized models
{[}15{]}. Most of this incremental work was indeed carried out in later
investigations in our sample.

\hypertarget{reproducibility}{%
\subsection{Reproducibility}\label{reproducibility}}

\label{sec:reproducibility}

@@

As discussed in Section \ref{sec:rationales}, most studies were premised
on the potential predictive value of localized models. As cautioned in
Section \ref{sec:performance}, not all experiments affirmed this
premise. Most of those that did involved using metric learning to
improve retrieval and sometimes iterative optimization of the metric and
of the localized model. Some studies showcased results using a variety
of such specializations {[}15,27,30{]}. However, because these advanced
implementations were mostly compared against their precursors or
commonplace alternatives, we cannot speak to their performance or any
trade-offs with respect to each other. This would be a worthy goal of
future work, though it might be limited by the lack of open-source code
or public implementations (Section \ref{sec:reproducibility}).

@@

Most deduplication was done automatically in Covidence. As full-text
review was done in Zotero, some additional duplicates were noticed and
merged.

When abstracts were not obtained by search or by Covidence, we attempted
to find them online using DOIs; when an abstract could not be found,
screening was based on the title alone. We decided to screen titles and
abstracts conservatively, rejecting only studies that were clearly
outside the scope of our review. The reasons for rejection were four, as
discussed in the main text: a. Exclude non clinical/non medical setting
b. Must be in English c.~Must be original study (not reviews, surveys,
opinion, news) d.~Exclude if search term clearly has different meaning
than intended (Two examples of (d) are the use of the term
``personalized model'' to refer in some cases to parameterized models
tuned to individual patient measurements and in others to
patient-centered models of care.) Studies that passed title/abstract
screen were exported to Zotero.

@@

\textbf{TODO: Place vertical separators between major columns.}
